-------
Peter Name

R227540N
-------------


```{r}
#Libraries
library(pacman)
pacman::p_load(dplyr,
               data.table,
               R.utils,
               igraph,
               reshape2)
```

Helper Functions

```{r}
set.seed(42)

# HELPER FUNCTIONS

# Simulate one trajectory of a Markov chain
simulate_mc <- function(P, start, n_steps = 50) {
  n_states <- nrow(P)
  traj <- integer(n_steps + 1)
  traj[1] <- start
  for (t in seq_len(n_steps)) {
    traj[t + 1] <- sample(n_states, 1, prob = P[traj[t], ])
  }
  traj
}

# Compute unconditional distribution at each step: pi_0 %*% P^n
unconditional_dist <- function(P, pi0, n_max = 60) {
  n_states <- ncol(P)
  probs <- matrix(0, nrow = n_max + 1, ncol = n_states)
  probs[1, ] <- pi0
  for (n in seq_len(n_max)) {
    probs[n + 1, ] <- probs[n, ] %*% P
  }
  probs
}

# Steady-state via left eigenvector (eigenvalue = 1)
steady_state <- function(P) {
  ev  <- eigen(t(P))
  idx <- which(abs(ev$values - 1) < 1e-9)[1]
  v   <- Re(ev$vectors[, idx])
  v / sum(v)
}

# GCD for period computation
gcd <- function(a, b) { if (b == 0) a else gcd(b, a %% b) }

# Period of state i (0-indexed internally): count step lengths of
# all paths i -> i and take their GCD.
period_state <- function(P, i, max_power = 60) {
  n  <- nrow(P)
  Pk <- diag(n)          # P^0
  d  <- 0
  for (k in seq_len(max_power)) {
    Pk <- Pk %*% P
    if (Pk[i, i] > 1e-12) d <- gcd(d, k)
  }
  if (d == 0) d <- Inf   # absorbing / never returns
  d
```

QUESTION 1A

```{r}
# QUESTION A1 part a
P1 <- matrix(c(
  1,0,0,0,0,
  0.5,0,0,0,0.5,
  0.2,0,0,0,0.8,
  0,0,1,0,0,
  0,0,0,1,0),
  nrow=5, byrow=TRUE)

states1 <- c("S1","S2","S3","S4","S5")
rownames(P1) <- colnames(P1) <- states1
mc1 <- new("markovchain", transitionMatrix=P1, states=states1)

# ── A1(a): Diagram & Classification ────────────────────────

par(mar=c(1,1,3,1))
plot(mc1,
     main="A1(a): 5-State Markov Chain",
     layout=layout_in_circle,
     vertex.size=45,
     vertex.label.cex=1.2,
     edge.label.cex=1,
     edge.arrow.size=0.5,
     vertex.color="dodgerblue4")

cat("\n=== A1(a) Classification ===\n")
cat("\nCommunicating classes:\n"); print(communicatingClasses(mc1))
cat("\nRecurrent classes:\n"); print(recurrentClasses(mc1))
cat("\nTransient classes:\n"); print(transientClasses(mc1))
cat("\nAbsorbing states:\n"); print(absorbingStates(mc1))

cat("\nPeriods:\n")
for (s in states1) {
  d <- state_period(P1, s, states1)
  print(paste(s, d))
}
```
```{r}
#QUESTION A1 part b
set.seed(42)
n_steps <- 30
state_num <- function(s) match(s, states1)

traj <- replicate(3, {
  s0 <- sample(states1,1)
  c(s0, rmarkovchain(n_steps, mc1, t0=s0))
}, simplify=FALSE)

par(mar=c(5,5,4,2))

plot(0:n_steps, sapply(traj[[1]], state_num),
     type="o", pch=16, lwd=2,
     col="blue", ylim=c(1,5),
     yaxt="n",  # 🔑 THIS FIXES THE DOUBLE AXIS
     xlab="Time Step", ylab="State",
     main="A1(b): Simulated Trajectories")

# Custom clean y-axis
axis(2, at=1:5, labels=states1, las=1)

lines(0:n_steps, sapply(traj[[2]], state_num),
      type="o", col="red", pch=16, lwd=2)

lines(0:n_steps, sapply(traj[[3]], state_num),
      type="o", col="darkgreen", pch=16, lwd=2)

legend("topright",
       legend=c("Trajectory 1","Trajectory 2","Trajectory 3"),
       col=c("blue","red","darkgreen"),
       lwd=2, pch=16, bty="n")

cat("\nComment: Chains eventually get absorbed into S1.\n")
```


```{r}
#Question A1 part c

cat("\n--- (c) Steady-State Probabilities ---\n")
pi1 <- steady_state(P1)
names(pi1) <- paste0("S", 1:5)
cat("Steady-state vector π:\n"); print(round(pi1, 6))

cat("\nInterpretation:\n")
cat("π(S1) + π(S4) = 1 (all probability absorbed into the two absorbing states).\n")
cat("π(S2) = π(S3) = π(S5) = 0 (transient states vanish at stationarity).\n")
cat("The exact split between S1 and S4 depends on starting distribution.\n")
cat("\nIs it ergodic? NO.\n")
cat("Ergodicity requires a single recurrent class that is aperiodic.\n")
cat("Here we have TWO recurrent classes + transient states => NOT ergodic.\n")

```



```{r}
# Question A1 part d
n_time <- 50
init <- rep(1/5,5)
prob_mat <- matrix(0, n_time+1, 5)
prob_mat[1,] <- init

for(t in 1:n_time){
  prob_mat[t+1,] <- prob_mat[t,] %*% P1
}

par(mar=c(5,5,4,5))
matplot(0:n_time, prob_mat,
        type="l", lwd=3, lty=1,
        col=rainbow(5),
        xlab="Time", ylab="Probability",
        main="A1(d): Probability Convergence",
        ylim=c(0,1))

legend("right", legend=states1,
       col=rainbow(5), lwd=3,
       inset=c(-0.2,0), xpd=TRUE, bty="n")

grid()

#The comment is below
```
The chain converges toward the steady-state distribution π = [1, 0, 0, 0, 0] within
approximately 50 to 60 steps. The convergence speed is governed by the smallest
exit probabilities from the transient states. The oscillating decay in states S3, S4,
and S5 reflects the periodic nature of the transient sub-group, where the probability
’leaks’ into the absorbing state S1 once every three steps on average.

```{r}
#QUESTION 2A part a
# a. Define the 7x7 transition matrix P2
P2 <- matrix(c(
  0,   1,   0,   0,   0,   0,   0,
  1,   0,   0,   0,   0,   0,   0,
  0,   0,   0,   0.4, 0.2, 0.2, 0.2,
  0,   0,   0,   0,   0.2, 0.4, 0.4,
  0.3, 0,   0,   0.1, 0.3, 0.1, 0.2,
  0,   0,   0,   0.2, 0.2, 0.3, 0.3,
  0,   0,   0,   0.5, 0.2, 0.2, 0.1
), nrow = 7, byrow = TRUE)

rownames(P2) <- colnames(P2) <- c("S1","S2","S3","S4","S5","S6","S7")

# 2. Build graph
g2 <- graph_from_adjacency_matrix(
  P2,
  mode     = "directed",
  weighted = TRUE,
  diag     = TRUE
)

# 3. FIX: Remove zero-weight edges using correct syntax
g2 <- delete_edges(g2, E(g2)[E(g2)$weight == 0])

# 4. Edge labels
E(g2)$label <- round(E(g2)$weight, 2)

# 5. Layout
lay2 <- matrix(c(
  -2,  1,   # S1
  -2, -1,   # S2
  0,   2,   # S3
  2,   1,   # S4
  0,   0,   # S5
  2, -1,   # S6
  0, -2    # S7
), ncol = 2, byrow = TRUE)

# 6. Vertex colours
vcols <- c("#E74C3C","#E74C3C","#F39C12","#3498DB","#3498DB","#3498DB","#3498DB")

# 7. Plot with enhanced visibility for recurrences (loops)
plot(g2,
     layout           = lay2,
     vertex.size      = 30,
     vertex.color     = vcols,
     vertex.label     = V(g2)$name,
     vertex.label.cex = 1.0,
     edge.label       = E(g2)$label,
     edge.label.cex   = 0.8,
     edge.curved      = 0.2,
     edge.arrow.size  = 0.3,    # Increased for better visibility
     edge.loop.angle  = -pi, # Adjusts the direction of self-loops
     # --- ADD THESE TWO LINES ---
     loop.size        = 1.5,   # Pushes the loop outside the vertex circle
     edge.arrow.width = 1.5,   # Makes the arrowhead bulky enough to see

     main             = "A2: 7-State Markov Chain")

legend("bottomleft",
       legend = c("Recurrent class {S1, S2} — period 2",
                  "Transient: S3",
                  "Recurrent class {S4, S5, S6, S7}"),
       fill   = c("#E74C3C","#F39C12","#3498DB"),
       bty    = "n",
       cex    = 0.8)

```
```{r}
#QUESTION A2 part b

cat("\nCommunicating classes:\n"); print(communicatingClasses(mc2))
cat("\nRecurrent classes:\n"); print(recurrentClasses(mc2))
cat("\nTransient classes:\n"); print(transientClasses(mc2))
cat("\nAbsorbing states:\n"); print(absorbingStates(mc2))

cat("\nPeriods:\n")
for (s in states2) {
  d <- state_period(P2, s, states2)
  print(paste(s, d))
}
```
```{r}
#QUESTION A2 part c
set.seed(2024)
n_steps2 <- 50
state_num2 <- function(s) match(s, states2)

traj2 <- replicate(2, {
  s0 <- sample(states2,1)
  c(s0, rmarkovchain(n_steps2, mc2, t0=s0))
}, simplify=FALSE)

par(mar=c(5,5,4,2))
plot(0:n_steps2, sapply(traj2[[1]], state_num2),
     type="o", col="blue", pch=16,
     ylim=c(1,7),
     xlab="Time", ylab="State",
     main="A2(c): Two Trajectories")

lines(0:n_steps2, sapply(traj2[[2]], state_num2),
      type="o", col="red", pch=16)

legend("topright", legend=c("Trajectory 1","Trajectory 2"),
       col=c("blue","red"), lwd=2, pch=16)
```
The plot reveals that the Markov Chain is periodic and reducible. Specifically, states 1 and 2 form a closed communicating class with a period of 2, as evidenced by the consistent oscillation between these two values over 50 time steps. The absence of transitions to states 3–7 suggests these states are either unreachable from the starting positions or belong to a separate, disjoint component of the state space.



```{r}
#QUESTION A2 part d 
# For the {S4,S5,S6,S7} sub-chain solve πQ = π
P2_sub <- P2[4:7, 4:7]
P2_sub_norm <- P2_sub / rowSums(P2_sub)   # already sums to 1 within sub-block
pi_sub <- steady_state(P2_sub_norm)
cat("Limiting probs within recurrent class {S4,S5,S6,S7}:\n")
names(pi_sub) <- paste0("S", 4:7)
print(round(pi_sub, 6))

cat("\n[1] RECURRENT CLASS ANALYSIS: {S1, S2}")
cat("\n----------------------------------------------------------")
cat("\n* Behavior:        Periodic (Period d = 2)")
cat("\n* Transition:      Direct oscillation between S1 and S2")
cat("\n* Limiting Prob:   0.5 (Time-average / Cesàro limit)")
cat("\n  (Note: A true limit does not exist due to oscillation.)")

cat("\n\n[2] TRANSIENT STATE ANALYSIS: {S3}")
cat("\n----------------------------------------------------------")
cat("\n* Status:          Transient")
cat("\n* Limiting Prob:   0.0")
cat("\n* Logic:           State is visited finitely often before")
cat("\n                   absorption into a recurrent class.")

cat("\n\n[3] ERGODICITY TEST")
cat("\n----------------------------------------------------------")
cat("\n* IS ERGODIC?      >>> NO <<<")
cat("\n* Reason 1:        REDUCIBILITY. The chain contains multiple")
cat("\n                   independent recurrent classes.")
cat("\n* Reason 2:        PERIODICITY. The {S1, S2} class violates")
cat("\n                   the aperiodicity requirement (d=2).")


```
```{r}
#Question A3 part b

P_day  <- matrix(c(0.4, 0.4, 0.2,
                   0.3, 0.4, 0.3,
                   0.0, 0.1, 0.9),
                 nrow = 3, byrow = TRUE)

P_peak <- matrix(c(0.1, 0.5, 0.4,
                   0.1, 0.3, 0.6,
                   0.0, 0.1, 0.9),
                 nrow = 3, byrow = TRUE)

traffic_states <- c("Light","Heavy","Jammed")
rownames(P_day)  <- colnames(P_day)  <- traffic_states
rownames(P_peak) <- colnames(P_peak) <- traffic_states

mat_power <- function(M, n) {
  result <- diag(nrow(M))
  for (i in seq_len(n)) result <- result %*% M
  result
}

#Analytical Distribution at 6 PM

pi0     <- c(1, 0, 0)
P_day9  <- mat_power(P_day,  9)
P_peak6 <- mat_power(P_peak, 6)
pi_4pm  <- pi0    %*% P_day9
pi_6pm  <- pi_4pm %*% P_peak6

cat("\n=== A3(a) Analytical Distribution ===\n")
cat("Distribution at 4 PM:\n");  print(round(pi_4pm, 4))
cat("Distribution at 6 PM:\n");  print(round(pi_6pm, 4))
cat(sprintf("P(Light  at 6 PM) = %.4f\n", pi_6pm[1]))
cat(sprintf("P(Heavy  at 6 PM) = %.4f\n", pi_6pm[2]))
cat(sprintf("P(Jammed at 6 PM) = %.4f\n", pi_6pm[3]))
```
```{r}
#QUSTION A3 part b
set.seed(123)
N <- 10000

norm_rows <- function(M) t(apply(M, 1, function(r) r / sum(r)))
P1n <- norm_rows(P_day)
P2n <- norm_rows(P_peak)

simulate_traffic <- function() {
  state <- 1
  for (t in 1:9) state <- sample(1:3, 1, prob = P1n[state, ])
  for (t in 1:6) state <- sample(1:3, 1, prob = P2n[state, ])
  state
}

final_states <- replicate(N, simulate_traffic())
emp_dist <- tabulate(final_states, nbins = 3) / N
names(emp_dist) <- traffic_states

#cat("\n=== A3(b) Monte Carlo Results (N = 10,000) ===\n")
#comparison <- rbind(Analytical = round(as.numeric(pi_6pm), 4),
 #                   Simulation = round(emp_dist, 4))
#colnames(comparison) <- traffic_states
#print(comparison)

# Comparison Bar Chart

par(mar = c(5, 5, 4, 3))
bp <- barplot(comparison,
              beside      = TRUE,
              col         = c("steelblue","firebrick"),
              border      = "white",
              main        = "A3(b): Analytical vs Simulated Distribution at 6 PM",
              ylab        = "Probability",
              xlab        = "Traffic State",
              ylim        = c(0, max(comparison) * 1.25),
              cex.main    = 1.2,
              cex.lab     = 1.1,
              cex.names   = 1.05)
legend("topright", legend = rownames(comparison),
       fill = c("steelblue","firebrick"),
       border = "white", bty = "n", cex = 0.95)

# Value labels on bars
text(bp, comparison + 0.012,
     labels = sprintf("%.3f", comparison),
     cex = 0.85, font = 2)

cat("\nComment: Simulation proportions closely match the analytical\n")
cat("probabilities. Jammed dominates at 6 PM due to peak-hour\n")
cat("transition probabilities pushing mass toward congestion.\n")

# Distribution of traffic
p <- ggplot(df_pm, aes(step, Prob, colour = State)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  geom_vline(xintercept = n_pm, linetype = "dashed", colour = "grey40") +
  annotate("text", x = n_pm + 0.3, y = 0.85, label = "4PM\n(matrix\nchanges)",
           hjust = 0, size = 3.2, colour = "grey30") +
  scale_x_continuous(breaks = all_steps,
                     labels = df_path$time) +
  scale_colour_manual(values = c("#27AE60","#E67E22","#E74C3C")) +
  labs(title = "A3: Distribution of Traffic State from 1PM to 6PM",
       x = "Time", y = "Probability") 
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)

```

```{r}
# QUESTION A3 part b

# Simulation
set.seed(123)
N <- 10000

norm_rows <- function(M) t(apply(M, 1, function(r) r / sum(r)))
P1n <- norm_rows(P_day)
P2n <- norm_rows(P_peak)

simulate_traffic <- function() {
  state <- 1
  for (t in 1:9) state <- sample(1:3, 1, prob = P1n[state, ])
  for (t in 1:6) state <- sample(1:3, 1, prob = P2n[state, ])
  state
}

final_states <- replicate(N, simulate_traffic())
emp_dist <- tabulate(final_states, nbins = 3) / N
names(emp_dist) <- traffic_states


# This matrix is necessary for the barplot to function
comparison <- rbind(Analytical = round(as.numeric(pi_6pm), 4),
                    Simulation = round(emp_dist, 4))
colnames(comparison) <- traffic_states
print(comparison)
# --- Plot 1: Comparison Bar Chart ---
par(mar = c(5, 5, 4, 3))
bp <- barplot(comparison,
              beside      = TRUE,
              col         = c("steelblue","firebrick"),
              border      = "white",
              main        = "A3(b): Analytical vs Simulated Distribution at 6 PM",
              ylab        = "Probability",
              xlab        = "Traffic State",
              ylim        = c(0, max(comparison) * 1.25),
              cex.main    = 1.2,
              cex.lab     = 1.1,
              cex.names   = 1.05)

legend("topright", legend = rownames(comparison),
       fill = c("steelblue","firebrick"),
       border = "white", bty = "n", cex = 0.95)

text(bp, comparison + 0.012,
     labels = sprintf("%.3f", comparison),
     cex = 0.85, font = 2)

# --- Plot 2: Distribution of traffic ---
p <- ggplot(df_pm, aes(step, Prob, colour = State)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  geom_vline(xintercept = n_pm, linetype = "dashed", colour = "grey40") +
  annotate("text", x = n_pm + 0.3, y = 0.85, label = "4PM\n(matrix\nchanges)",
           hjust = 0, size = 3.2, colour = "grey30") +
  scale_x_continuous(breaks = all_steps,
                     labels = df_path$time) +
  scale_colour_manual(values = c("#27AE60","#E67E22","#E74C3C")) +
  labs(title = "A3: Distribution of Traffic State from 1PM to 6PM",
       x = "Time", y = "Probability") + 
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)
```

